### 2018-05-25 submission/manuscript.Rmd
complete re-write of the abstract

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and polymerase can change the bacterial community and introduce both error and bias to the final obtained results.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities. It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and chimera prevalence correlation with the number of OTUs.

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).

**Conclusions.** Although the variation in the number of OTUs observed in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for this difference. In mock community samples, the divergence in the number of OTUs between the polymerases which can be explained by chimera prevalence suggests that this is the main reason for differences between the polymerases. Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.

>>>>

**Background.** It is challenging to compare 16S rRNA gene sequencing data across studies and one of the reasons for this is due to error. There are many different places throughout the workflow where error can be introduced into the pipeline. Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error into this workflow because of less investigation into this area. By varying both cycle number and polymerase we assess the changes that occur to the bacterial community in human fecal and mock community samples.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities. It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).

**Conclusions.** Although the variation in the number of OTUs in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for our observed differences. In mock community samples, the strong correlation between chimera prevalence and the number of OTUs suggests that this is the main reason for differences between the polymerases. Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles in 16S rRNA gene sequencing studies.

### 2018-05-25 submission/manuscript.tex
complete re-write of the abstract

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Although the number of cycles and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and polymerase can change the bacterial community and
introduce both error and bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, chimera prevalence, and chimera prevalence
correlation with the number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). Variations in the
number of OTUs detected by DNA polymerase was observable as early as 20
cycles in mock communities (P-value = 0.002). There also was a large
range in the number of OTUs by HiFi DNA polymerase at 35 cycles
(Accuprime = 15 - 20 versus Phusion = 14 - 73). Chimera prevalence in
mock communities varied by polymerase with differences being most
notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs with Platinum having the highest (R\^{}2 = 0.974) and
Kappa having the worse (R\^{}2 = 0.478).

\textbf{Conclusions.} Although the variation in the number of OTUs
observed in fecal samples could be due to certain polymerases capturing
the biological variability better than others, this is unlikely to be
the main reason for this difference. In mock community samples, the
divergence in the number of OTUs between the polymerases which can be
explained by chimera prevalence suggests that this is the main reason
for differences between the polymerases. Ultimately, this variation
makes comparison across studies difficult and care should be exercised
when choosing the polymerase and number of cycles to be used in 16S rRNA
gene sequencing studies.

>>>>

\textbf{Background.} It is challenging to compare 16S rRNA gene
sequencing data across studies and one of the reasons for this is due to
error. There are many different places throughout the workflow where
error can be introduced into the pipeline. Here, we focus on studying
how the number of cycles and high fidelity (HiFi) DNA polymerase
introduce error into this workflow because of less investigation into
this area. By varying both cycle number and polymerase we assess the
changes that occur to the bacterial community in human fecal and mock
community samples.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, chimera prevalence, and the correlation
between chimera prevalence and number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variations in the number of OTUs detected by
DNA polymerase was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs by HiFi DNA
polymerase at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\^{}2 = 0.974) and Kappa having the worse (R\^{}2 = 0.478).

\textbf{Conclusions.} Although the variation in the number of OTUs in
fecal samples could be due to certain polymerases capturing the
biological variability better than others, this is unlikely to be the
main reason for our observed differences. In mock community samples, the
strong correlation between chimera prevalence and the number of OTUs
suggests that this is the main reason for differences between the
polymerases. Ultimately, this variation makes comparison across studies
difficult and care should be exercised when choosing the polymerase and
number of cycles in 16S rRNA gene sequencing studies.
